Robert C. Dunnell’s Systematics in Prehistory at 50

### 2021-07-18 analysis/paper/paper.Rmd
update paper

While the influence of *Systematics* has extended well beyond Dunnell's own primary study area of the US Southeast, there are some notable limits to its contributions. In our data we see only work in an East or Southeast Asian language, a book chapter in Japanese surveying the literature on cultural phylogenetics, indicating limited attention to *Systematics* from researchers in the Eastern hemisphere. Could it be that cultural differences in reasoning styles [@henrich2010weirdest] result in a diminished relevance for the philosophical content of *Systematics* outside of Western academic communities? Or perhaps it is simply that book is quite difficult to read, a detail noted by the reviewers when it was first published ('confusion followed by grudging agreement' is how Spaulding [@spaulding1974] described his reaction to the book), and acknowledged by Dunnell in his forward to the 2002 edition. The low readability of the book likely limited the accessibility of its contents for readers whose first language is not English. A second limitation, or maybe better described as an unrealised contribution, of *Systematics* is the absence of a distinct research area and community of archaeological systemics. While biology has a Society of Systematic Biologists, and journals such as _Systematic Biology_, _Systematics and Evolution_, and _Systematics and Biodiversity_, archaeology has no equivalent focal points for discussions of systematics that transcends specific geographical and chronological concerns. *Systematics* was not a work that provided a universal motor force for moving archaeologists together at scale to tackle questions about the formation of units of measurement and classification. Instead these discussions happen deep within discrete research communities, constrained by the culture and norms of those groups.

Dunnell intended *Systematics* to establish a distinctive form of scientific archaeology, and it faced overwhelming competition from the program advocated by New Archaeology, which dominated the literature in the decade after the publication of *Systematics*. What makes *Systematics* a classic contribution is that is provides archaeologists with the master key to release themselves from the 'prison of de Mortillet' [@shea2016stone p. xvii], namely the inherited and entrenched habits of research traditions (Gabriel de Mortillet, 1821–1898, was an archaeologist who published the first widely used classification of the Paleolithic, much of which remains in use today). *Systematics* endures as a striking and precocious provocation to archaeologists to be transparent, rigorous, precise, and deliberate about how we divide up and aggregate material culture and the measurements we take from it to describe and explain the human experience in the past. With archaeologists increasingly pursuing ambitious questions about evolutionary processes that require large scale syntheses of disparate datasets, the message of *Systematics* will only become more relevant.

>>>>

While the influence of *Systematics* has extended well beyond Dunnell's own primary study area of the US Southeast, there are some notable limits to its contributions. In our data we see only work in an East or Southeast Asian language, a book chapter in Japanese surveying the literature on cultural phylogenetics, indicating limited attention to *Systematics* from researchers in the Eastern hemisphere. Could it be that cultural differences in reasoning styles [@henrich2010weirdest] result in a diminished relevance for the philosophical content of *Systematics* outside of Western academic communities? Or perhaps it is simply that book is quite difficult to read, a detail noted by the reviewers when it was first published ('confusion followed by grudging agreement' is how Spaulding [@spaulding1974] described his reaction to the book), and acknowledged by Dunnell in his forward to the 2002 edition. The low readability of the book likely limited the accessibility of its contents for readers whose first language is not English. A second limitation, or maybe better described as an unrealised contribution, of *Systematics* is the absence of a distinct research area and community of archaeological systemics. While biology has a Society of Systematic Biologists, and journals such as _Systematic Biology_, _Systematics and Evolution_, and _Systematics and Biodiversity_, archaeology has no equivalent focal points for discussions of systematics that transcends specific geographical and chronological concerns. *Systematics* was not a work that provided a universal motor force for moving archaeologists together at scale to tackle questions about the formation of units of measurement and classification. Instead these discussions typically happen deep within discrete and disconnected archaeological research communities, constrained by the culture and norms of those groups.

Dunnell intended *Systematics* to establish a distinctive form of scientific archaeology, and it faced overwhelming competition from the program advocated by New Archaeology. The approaches of the New Archaeologists dominated the literature in the decade after the publication of *Systematics*, thorouhgly eclipsed it as a discipline-defining text. Nevertheless, what makes *Systematics* a classic contribution is that is provides archaeologists with the master key to release themselves from the 'prison of de Mortillet' [@shea2016stone p. xvii], namely the inherited and entrenched analytical habits of research traditions dominated by prestige economies (Gabriel de Mortillet, 1821–1898, was an archaeologist who published the first widely used classification of the Paleolithic, much of which remains in use today). *Systematics* endures as a striking and precocious provocation to archaeologists to be transparent, rigorous, precise, and deliberate about how we divide up and aggregate material culture and the measurements we take from it to describe and explain the human experience in the past. With archaeologists increasingly pursuing ambitious questions about evolutionary processes that require large scale syntheses of disparate datasets, the message of *Systematics* will only become more relevant.
